---
title: "Cyclistic"
author: "Rion Sutradhar"
date: "2024-06-14"
output: html_document
---

# All necessary libraries and installed and initialised

```{r Libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
library(scales)
library(dplyr)
library(gridExtra)
```

# *Set the directory to the location where all the files are present*

```{r Set Working Directory, echo=TRUE}
setwd('C:/Users/rions/Desktop/My Projects/Cyclistic bike share analysis')
```

# *Load all the datasets of individual months for 2023*

```{r Datasets}
jandf <- read.csv("./Data/202301-divvy-tripdata.csv")
febdf <- read.csv("./Data/202302-divvy-tripdata.csv")
mardf <- read.csv("./Data/202303-divvy-tripdata.csv")
aprdf <- read.csv("./Data/202304-divvy-tripdata.csv")
maydf <- read.csv("./Data/202305-divvy-tripdata.csv")
jundf <- read.csv("./Data/202306-divvy-tripdata.csv")
juldf <- read.csv("./Data/202307-divvy-tripdata.csv")
augdf <- read.csv("./Data/202308-divvy-tripdata.csv")
sepdf <- read.csv("./Data/202309-divvy-tripdata.csv")
octdf <- read.csv("./Data/202310-divvy-tripdata.csv")
novdf <- read.csv("./Data/202311-divvy-tripdata.csv")
decdf <- read.csv("./Data/202312-divvy-tripdata.csv")
```

# *Merge to single data frame*

```{r Merge datasets}
bikes_data <-rbind(jandf, febdf, mardf, aprdf, maydf, jundf, juldf, 
                   augdf, sepdf, octdf, novdf, decdf)

```

# *Remove all empty Rows and Columns*

```{r Empty R&C, echo=TRUE}
bikes_data <- remove_empty(bikes_data, which = c('cols', 'rows'))
summary(bikes_data)
```

# *Drop all null values*

```{r NULL Values, echo=TRUE}
bikes_data <- bikes_data %>%
  drop_na()
summary(bikes_data)
```

# *Parse strings and convert them to Date/Time*

```{r Convert to date & time}
bikes_data$started_at <- ymd_hms(bikes_data$started_at)
bikes_data$ended_at <- ymd_hms(bikes_data$ended_at)
```

# *Calculate trip duration of all instances in minutes*

```{r Trip duration}
bikes_data <- bikes_data %>%
  mutate(trip_duration=as.numeric(difftime(ended_at, started_at, units="mins")))
```

# *Filter out negative trip duration*

```{r (-ve) trip remove}
bikes_data <- bikes_data %>%
  filter(trip_duration > 0)
```

# *Calculate day of week and hour of day*

```{r Week & hour calculation, echo=TRUE}
bikes_data <- bikes_data %>%
  mutate(day_of_week = wday(started_at, label = TRUE, abbr = FALSE), hour_of_day = hour(started_at))
head(bikes_data)
```

# *Trip duration summary of mean, median, max and min*

```{r Statistics, echo=FALSE}
trip_duration_summary <- bikes_data %>%
  group_by(member_casual) %>%
  summarise(mean_trip_duration = mean(trip_duration),
            median_trip_duration = median(trip_duration),
            max_trip_duration = as.numeric(max(trip_duration)),
            min_trip_duration = as.numeric(min(trip_duration)))
print(trip_duration_summary)
```

# *Day of week summary to count the number of times bike has been rented in a day arranged in ascending order*

```{r No. of Trips per day, echo=TRUE, message=FALSE}
day_of_week_summary <- bikes_data %>%
  group_by(member_casual, day_of_week) %>%
  summarise(trip_count = n()) %>%
  arrange(member_casual, day_of_week)
print(day_of_week_summary)
```

# *Hour of day summary to count the number of times bike has been rented in an hour arranged in ascending order*

```{r No. of Trips per hour, echo=TRUE, message=FALSE}
hour_of_day_summary <- bikes_data %>%
  group_by(member_casual, hour_of_day) %>%
  summarise(trip_count = n()) %>%
  arrange(member_casual, hour_of_day)
print(hour_of_day_summary, n=100)
```

```{r Mean trip Graph, include=FALSE}
mean_plot <- ggplot(trip_duration_summary, aes(x = member_casual, y = mean_trip_duration, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Trip Duration by User Type", y = "Mean Trip Duration (minutes)", x = "User Type") +
  theme_minimal()
```

```{r Median trip Graph, include=FALSE}
median_plot <- ggplot(trip_duration_summary, aes(x = member_casual, y = median_trip_duration, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Median Trip Duration by User Type", y = "Median Trip Duration (minutes)", x = "User Type") +
  theme_minimal()
```

# *Bar plot for mean and median trip duration*

```{r Grid View, echo=FALSE}
grid.arrange(mean_plot, median_plot)
```

# *Plot for trips by day of week*

```{r Trips per day in a week plot, echo=FALSE}
ggplot(day_of_week_summary, aes(x = day_of_week, y = trip_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels=comma) +
  labs(title = "Trips by Day of Week", y = "Number of Trips", x = "Day of Week")
```

# *Plot trips by hour of day*

```{r Trips per hour of day graph, echo=FALSE}
ggplot(hour_of_day_summary, aes(x = hour_of_day, y = trip_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels=comma) +
  labs(title = "Trips by Hour of Day", y = "Number of Trips", x = "Hour of Day")
```